---
title: "PML"
author: "Ivo Pinheiro"
date: "`r Sys.Date()`"
output: html_document
---

## Practical Machine Learning - Course Project

**To predict the manner in which participants performed barbell lifts using sensor data, that was the task at hand. To accomplish this, I cleaned and analysed data to built a model that predicts the value of a target variable ($classe) based on input variables (53 features).**

```{r}
# Libraries
set.seed(123)
library(ggplot2)
library(randomForest)
library(caret)
```

### Data preprocessing

I started by assigning NA to any missing value in the data, and then I removed any column that had over 90% NA. As it didn't make sense to include the first six columns, these were removed. Lastly, the target variable ($classe) was converted into a factor variable.

**These steps reduced the number of columns from 160 to 54, and ensured that our data was now ready for analysis and building our model.**

```{r}
# Importing
url_training <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_testing <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# Reading
training_data <- read.csv(url_training)
testing_data <- read.csv(url_testing)
# Cleaning training set
dim(training_data)
training_data[training_data == ""] <- NA
proportion_NA_train <- colSums(is.na(training_data))/nrow(training_data)
training_data1 <- training_data[, proportion_NA_train < 0.9]
training_data2 <- training_data1[, -c(1:6)]
training_data2$classe <- as.factor(training_data2$classe)
```

### Model building and evaluation 

Since this is a classification problem and we already know the target variable ($classe), the model that made the most sense to me was random forests. I trained the model on 70% of the training set and then I tested it on the other 30%.

```{r}
# Model with training set
in_Train <- createDataPartition(y=training_data2$classe, p=0.7, list=FALSE)
trainingset <- training_data2[in_Train, ]
testingset <- training_data2[-in_Train, ]
model_1 <- randomForest(classe ~ ., data = trainingset)
print(model_1)
````

Happy with the results, I applied the model to the testing set.

```{r}
# Cleaning test set
dim(testing_data)
testing_data[testing_data == ""] <- NA
proportion_NA_test <- colSums(is.na(testing_data))/nrow(testing_data)
testing_data1 <- testing_data[, proportion_NA_test < 0.9]
testing_data2 <- testing_data1[, -c(1:6)]
# Applying model to test set
testset_predictions <- predict(model_1, newdata = testing_data2)
print(testset_predictions)
```

**So what is the percentage of the target variable in the test set that are correctly classified by the model?**

```{r}
predictions <- predict(model_1, testingset)
confusionMatrix(predictions, testingset$classe)
```

**According to the confusion matrix, the model correctly classified 99.85% of the instances overall.**

**Given the high accuracy, the expected out-of-sample error is likely to be very low (1−0.9985=0.0015). **

**Thus, the expected out-of-sample error is approximately 0.15%.**


## Plots

```{r}
varImpPlot(model_1)

ggplot(training_data2, aes(x = classe)) + 
        geom_bar(fill = "white", color = "grey") + 
        labs(title = "How many of each $classe in the training set", x = NULL, y = NULL)

````



